---
title: "Using the Bayesian Adaptive Sampling (BAS) Package for Bayesian Model Averaging"
author: "Merlise A Clyde"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Bayesian Adaptive Sampling (BAS) Package for Bayesian Model Averaging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup}
require(GGally)
```

The `BAS` package provides easy to use functions to implement Bayesian Model Averaging in linear models and  generalized linear models.
Prior distributions on coefficients are
based on Zellner's g-prior or mixtures of g-priors, such as the 
Zellner-Siow Cauchy prior or mixtures of g-priors from 
[Liang et al (2008)](http://dx.doi.org/10.1198/016214507000001337)
for linear models, as well as other options including AIC, BIC, RIC and Empirical Bayes methods.  Extensions to Generalized Linear Models are based on the 
mixtures of g-priors in GLMs of
[Li and Clyde (2015)](http://arxiv.org/abs/1503.06913) using an integrated 
Laplace approximation.

`BAS` uses an adaptive sampling algorithm  to sample without replacement from the space of models or MCMC sampling which is recommended for sampling problems with a large number of predictors. See [Clyde, Littman & Ghosh](http://dx.doi.org/10.1198/jcgs.2010.09049) for more details for the sampling algorithms.

## Installing BAS

The stable version can be installed easily in the `R` console like any other package:

```r
install.packages('BAS')
```

On the other hand, I welcome everyone to use the most recent version
of the package with quick-fixes, new features and probably new
bugs.  To get the latest
development version from [GitHub](https://github.com/merliseclyde),
use the `devtools` package from
[CRAN](https://cran.r-project.org/package=devtools) and enter in `R`:

```r
# devtools::install_github('merliseclyde/BAS')
```

As the package does depend on BLAS and LAPACK, installing from GitHub will require that you have FORTRAN and C compilers on your system.  

## Demo 

We will use the UScrime data to illustrate some of the commands and functionality.

```{r data}
library(MASS)
data(UScrime)
```

Following other analyses, we will go ahead and log transform all of the variables except column 2, which is the indicator variable of the state being a southern state.

```{r transform}
UScrime[,-2] = log(UScrime[,-2])
```

To get started, we will use `BAS` with the Zellner-Siow Cauchy prior on the coefficients.

``` {r bas}
library(BAS)
crime.ZS =  bas.lm(y ~ ., 
                   data=UScrime,
                   prior="ZS-null",
                   modelprior=uniform(), initprobs="eplogp") 
```

`BAS` uses a model formula similar toe `lm` to specify the full model with all of the potential predictors.  Here we are using the shorthand `.` to indicate that all remaining variables in the data frame will be included.  `BAS` require a data frame as the input for the `data` argument.   Different prior distributions on the regression coefficients may be specified using the `prior` argument, and include

  * "BIC"
  * "AIC
  * "hyper-g"
  * "hyper-g-n"
  * "EB-local"

By default, `BAS` will try to enumerate all models, in this case $2^15$.  The prior distribution over the models is a `uniform()` distribution which assigns equal probabilities to all models.  The last optional argument `initprobs = eplogp` provides a way to initialize the sampling algorithm.

## Plots

Some graphical summaries of the output may be obtained by the `plot` function

```{r, fig.show='hold'}
plot(crime.ZS, ask=F)

```

which produces a panel of four plots.  The first is a plot of residuals and fitted values under Bayesian Model Averaging.  Ideally,  of our model assumptions hold, we will not see outliers or non-constant variance. The second plot shows the cumulative probabilty of the models in the order that they are sampled.  This plot indicates that the cumulative probability is leveling off as each additional model adds only a small increment to the cumulative probability, which earlier, there are larger jumps corresponding to sampling high probabilty models.  The third plot shows the dimension of each model (the number of regression coefficients includeing the intercept) versus the log of the marginal likelihood of the model.    The last plot shows the marginal posterior inclusion probabilities (pip) for each of the covariates, with marginal pips greater than 0.5 shown in red.  The variables with pip > 0.5 correspond to what is known as the median probability model.  Variables with high inclusion probabilties are generally important for explaining the data or prediction, but marginal inclusion probabilities may be small if there are extreme correlations among predictors, similar to how p-values may be large in the presence of mullticollinearity.

Individual plots may be obtained using the `which` option.
```{r pip, fig.width=5, fig.height=5}
plot(crime.ZS, which = 4, ask=FALSE, caption="", sub.caption="")
```

`BAS` has `print` and `summary` methods defined for objects of class `bas`.  Typing the objects name
```{r print}
crime.ZS
```

returns a summary of the marginal inclusion probabilities, while the `summary` function
provides
```{r summary}
summary(crime.ZS)
```
This lists the top 5 models (in terms of posterior probability)  with the zero-one indicators for variable inclusion.  The other columns in the summary are the Bayes factor of each model to the hiest probbility model (hence its Bayes factor is 1), the posterior probabilites of the models, the ordinary $R^2$ of the models, the dimension of the models (number of coeeficients includeing the intercept)  and the log marginal likelihood usnder the selected prior distribution. 

## Visualization of the Model Space  

To see beyond the first five models, we can represent the collection of the models via an `image` plot.  By default this shows the top 20 models.

```{r image, fig.width=5, fig.height=5}
image(crime.ZS, rotate=F)
```

This image has rows that correspond to each of the variables and intercept, 
with labels for the variables on the y-axis.  The x-axis corresponds to the possible models. These are sorted by their posterior probability from best at the left
to worst at the right with the rank on the top x-axis. 

Each column represents one of the 16 models.  The variables
that are excluded in a model are shown in black for each column, while
the variables that are included are colored, with the color related to
the log posterior probability. 
The color of each column is proportional to the log of the posterior probabilities (the lower x-axis) of that model. Models that are the same color have similar log posterior probabilities  which allows us to view models that are clustered together that have marginal likelihoods where the  differences are not "worth a bare mention".

This plot indicates that the police expenditure in the two years do not enter the model together, and is an indication of the high correlation between the two variables.

## Posterior Distributions of Coefficients

To examine the marginal distributions of the two coefficients for the police expenditures, we can estract their coefficents and then plot the posterior distributions averaging over all of the models.

```{r plot}
coef.ZS = coef(crime.ZS)
plot(coef.ZS, subset=c(5:6),  ask=F)
```

The vertical bar represents the posterior probability that the coefficient is 0 while
the  bell shaped curve represents the density of plausible values from all the models where the coefficient is non-zero.   This is scaled so that the height of the density for non-zero values is the probability that the coefficient is non-zero.  

Omitting the `subset` argument provides all of the marginal distributions

```{r coefall}

plot(coef.ZS, ask=FALSE)
```


To obtain credible interals for coefficients, `BAS` includes a `confint` method to create Highest Posterior Density intervals from the summaries from `coef`.
```{r confint-coef}

confint(coef.ZS)
```
where the third column is the posterior mean.
This uses Monte Carlo sampling to draw from the mixture model over coefficient where models are sampled based on their posterior probabilities.

We can also plot these via
```{r plot-confint}
plot(confint(coef.ZS, parm=2:16))
```
using the `parm` argument to select which coefficients to plot (the intercept is `parm=1`).


## Prediction

`BAS` has methods defined to return fitted values, `fitted`, using the observed design matrix and predictions at either the observed data or potentially new values, `predict`, as with `lm`.  

```{r choice of estimator}
muhat.BMA = fitted(crime.ZS, estimator="BMA")
BMA  = predict(crime.ZS, estimator="BMA")

# predict has additional slots for fitted values under BMA, predictions under each model
names(BMA)
```

Plotting the two sets of fitted values,
```{r, fig.width=5, fig.height=5}
par(mar=c(9, 9, 3, 3))
plot(muhat.BMA, BMA$fit, 
     pch=16, 
     xlab=expression(hat(mu[i])), ylab=expression(hat(Y[i])))
abline(0,1)
```

we see that they are in perfect agreement.  That is always the case as the posterior mean for the regression mean function at a point $x$ is the expected posterior  predictive value for $Y$ at $x$.   This is true not only for estimators such as BMA, but the expected values under model selection.   

### Inference with model selection ###

In addition to using BMA, we can use the posterior means under model selection.  This corresponds to a decision rule that combines estimation and selection.  `BAS` currently implements the following options


**highest probability model:**

```{r HPM}
HPM = predict(crime.ZS, estimator="HPM")

# show the indices of variables in the best model where 0 is the intercept
HPM$bestmodel
```

A little more interpretable version with names:
```{r}
(crime.ZS$namesx[HPM$bestmodel +1])[-1]
```

**median probability model:**
```{r MPM}
MPM = predict(crime.ZS, estimator="MPM")
(crime.ZS$namesx[attr(MPM$fit, 'model') +1])[-1]
```


Note that we can also extract the best model from the attribute in the fitted values as well.

**best predictive model:**

This is the model that is closest to BMA predictions under squared error loss.
```{r BPM}
BPM = predict(crime.ZS, estimator="BPM")
(crime.ZS$namesx[attr(BPM$fit, 'model') +1])[-1]
```

Let's see how they compare:

```{r}
library(GGally)
ggpairs(data.frame(HPM = as.vector(HPM$fit),  #this used predict so we need to extract fitted values
                   MPM = as.vector(MPM$fit),  # this used fitted
                   BPM = as.vector(BPM$fit),  # this used fitted
                   BMA = as.vector(BMA$fit))) # this used predict
```


Using the `se.fit = TRUE` option with `predict` we can also calculate standard deviations for prediction or for the mean and use this as imput for the `confint` function for the prediction object.

```{r se}
BPM = predict(crime.ZS, estimator="BPM", se.fit=TRUE)
crime.conf.fit = confint(BPM, parm="mean")
crime.conf.pred = confint(BPM, parm="pred")
cbind(crime.conf.fit, crime.conf.pred)
plot(crime.conf.fit)
```

## Beyond Enumeration

Many problems are too large to enumerate all possible models.  In such cases we may use the `method=MCMC` option to sample models using Markov Chain Monte Carlo sampling to sample models based on their posterior probabilities.

```{r MCMC}
crime.ZS =  bas.lm(y ~ ., 
                   data=UScrime,
                   prior="ZS-null",
                   modelprior=uniform(),
                   method = "MCMC") 
```

This will run the MCMC sampler until the number of uniques sampled models exceeds `n.models` which is $2^p$ (if $p < 19$) by default  or until `MCMC.iterations` has been exceeded, where `MCMC.iterations = n.models*2` by default.

### Estimates of Marginal Posterior Inclusion Probabilities (pip) ###

With MCMC sampling there are two estimates of the marginal inclusion probabilities:  `object$probne0` which are obtained by using the renormalized posterior odds from sampled models to estimate probabilities and the estimates based on Monte Carlo frequencies `object$probs.MCMC`.  These should be in close agreement if the MCMC sampler has run for enough iterations.

`BAS` includes a diagnostic function to compare the two sets of estimates of posterior inclusion probabilities and posterior model probabilities
```{r diagnostics}
diagnostics(crime.ZS, type="pip",  pch=16)
diagnostics(crime.ZS, type="model",  pch=16)
```

In the left hand plot of pips, each point represents one posterior inclusion probability for the 15 variables estimated under the two methods. The two estimators are in pretty close agreement.  The plot of the model probabilites suggests that we should use more `MCMC.iterations` if we want more accurate estimates of the posterior model probabilities.

```{r biggerMCMC}
crime.ZS =  bas.lm(y ~ ., 
                   data=UScrime,
                   prior="ZS-null",
                   modelprior=uniform(),
                   method = "MCMC", MCMC.iterations = 10^6)  

diagnostics(crime.ZS, type="model", pch=16)
```

## Summary

`BAS` includes other prior distributions on coefficients and models, as well as `bas.glm` for fitting Generalized Linear Models.  Some of the syntax for bas.glm and bas.lm have not converged, particularly how some of the priors are represented, so do please see the documentation for more features and details until this is updated!
